# Theoretical Background

This chapter sets out the concepts the thesis relies on: landscape connectivity and its modeling, the movement ecology of the wild boar, and the epidemiology of African Swine Fever (ASF), which together make connectivity a key factor in disease management.

## Ecological Connectivity in Fragmented Landscapes

Landscape connectivity describes the degree to which a landscape facilitates or impedes the movement of organisms among habitat patches. @taylor1993 formalised the concept as a vital structural element of landscapes, and it is now standard to separate two dimensions. Structural connectivity refers to the physical arrangement of habitat in space and can be read from land cover maps alone. Functional connectivity reflects the behavioural response of a specific species to that structure [@urbina2023; @tischendorf2003], so two structurally identical landscapes can differ greatly depending on whether the target species treats roads, rivers, or fields as permeable or as barriers. A resistance surface calibrated for one species therefore cannot simply be reused for another. This distinction matters in the Swiss Mittelland, which looks connected on a coarse map but is functionally fragmented for large mammals by dense road and rail networks, urban development, and intensive agriculture, leading to reduced gene flow and limited recolonisation for species such as red deer and wild boar [@fischer2024; @urbina2023].

## Modeling Functional Connectivity

The following section provides a theoretical explanation of all the modeling methods for functional connectivity used in this study.

### Resistance Surfaces and the Integrated Step-Selection Function

The resistance surface is the core spatial input to all functional connectivity models. It is a raster in which each cell carries a value representing the relative cost of movement, with high-quality habitat assigned low resistance and barriers such as motorways assigned very high or infinite resistance. Resistance can be assigned from expert judgement or, more defensibly, estimated empirically from GPS tracking. @urbina2023 derived resistance for red deer from over 85,000 GPS fixes of 15 individuals across the Greater Geneva agglomeration at 5-metre resolution, and @fischer2024 applied a comparable empirical framework to the Swiss Mittelland. Finer resolution resolves the narrow infrastructure that constrains large-mammal movement but multiplies the cell count and the computational cost.

Where GPS data are available, the integrated step-selection function (iSSF) provides a principled basis for this estimation [@avgar2016]. The iSSF reconstructs movement as a sequence of steps, each with a step length and turning angle, and compares every observed step against random alternatives drawn from the same movement kernel. Fitting a conditional logistic regression of the form $w(x) = \exp\!\left(\sum_k \hat{\beta}_k x_k\right)$ yields selection coefficients $\hat{\beta}_k$ that express habitat preference while holding movement capacity and availability constant. This separates genuine preference from mere abundance: an animal seen often in forest may prefer forest, or forest may simply be plentiful where it started, and the conditional design distinguishes the two. To turn the coefficients into resistance on the scale required by corridor models, @keeley2016 proposed the transformation $R = \exp\!\left(-c \cdot \hat{\beta}_k x_k\right)$, where the sign is reversed so that strongly selected cells receive low resistance. A single tracking dataset therefore yields a resistance surface that is both species-specific and statistically grounded.

```{r}
#| label: fig-issf-schematic
#| echo: false
#| fig-width: 9
#| fig-height: 4.2
#| fig-cap: "Schematic of the integrated step-selection function and the resulting resistance surface. Left: at each observed location the animal's realised step (solid arrow) is compared against random available steps (dashed arrows) drawn from its movement kernel, so the model learns which features are selected. Right: the fitted coefficients are translated into a resistance raster, where preferred habitat receives low resistance (green) and barriers such as roads receive high resistance (red)."

op <- par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1))

# Panel A: used vs. available steps
plot(0, 0, type = "n", xlim = c(0, 10), ylim = c(0, 10), axes = FALSE,
     xlab = "", ylab = "", main = "iSSF: used vs. available steps")
rect(5.5, 5.5, 10, 10, col = "#cdebd0", border = NA); text(7.7, 9.5, "forest", col = "#2d6a4f", cex = 0.8)
rect(0, 0, 4.5, 4, col = "#f1e2c4", border = NA);     text(2.2, 0.5, "open field", col = "#9c7a30", cex = 0.8)
x0 <- 4; y0 <- 4
set.seed(1)
for (a in seq(0, 2 * pi, length.out = 8)[-8])
  arrows(x0, y0, x0 + 3 * cos(a), y0 + 3 * sin(a), length = 0.06, col = "grey65", lty = 2)
arrows(x0, y0, 6.6, 6.6, length = 0.10, col = "#1b4332", lwd = 2.5)
points(x0, y0, pch = 19, cex = 1.3)
text(x0 - 0.3, y0 - 0.6, "current fix", cex = 0.7, adj = 1)
text(6.8, 6.3, "observed step", cex = 0.7, adj = 0, col = "#1b4332")

# Panel B: resistance raster
plot(0, 0, type = "n", xlim = c(0, 10), ylim = c(0, 10), axes = FALSE,
     xlab = "", ylab = "", main = "Resistance surface")
pal <- colorRampPalette(c("#2d6a4f", "#fdfdbf", "#9b2226"))(100)
for (i in 0:9) for (j in 0:9) {
  val <- 0.4
  if (i >= 5 && j >= 5) val <- 0.1
  if (i < 4 && j < 4) val <- 0.55
  rect(i, j, i + 1, j + 1, col = pal[round(val * 99) + 1], border = "white", lwd = 0.5)
}
for (i in 0:9) rect(i, 4, i + 1, 5, col = "#9b2226", border = "white", lwd = 0.5)
text(5, 4.5, "road (high R)", col = "white", cex = 0.7)
legend("bottomleft", fill = c("#2d6a4f", "#fdfdbf", "#9b2226"),
       legend = c("low", "med", "high"), bty = "n", cex = 0.7, title = "resistance")
par(op)
```

### Least-Cost Paths and Circuit Theory

Least-cost path (LCP) analysis was among the first methods to use resistance surfaces for corridor mapping [@adriaensen2003]. It identifies the single route between two patches that minimises cumulative resistance. It is computationally cheap and widely used, but it assumes movement along one optimal route and says nothing about whether alternative routes exist or how redundant the surrounding landscape is [@mcrae2008].

Circuit theory addresses that limitation by treating the resistance surface as an electrical network in which current flows from a source to a ground through all available pathways at once [@mcrae2008]. @mcrae2008 showed that the current through any cell is proportional to the expected number of times a random walker would pass through it, giving the method a process-based interpretation that LCP lacks. High current density marks heavily used corridors, and points where the landscape forces nearly all movement through a single passage are called pinch points, which are the highest-priority targets for intervention because losing a small area there disproportionately reduces connectivity. Because effective resistance falls as parallel routes are added, the method also captures redundancy, and the ratio of least-cost to resistance distance measures how much a single bottleneck dominates. The Circuitscape software solves networks of millions of cells, making it applicable at landscape scale. @urbina2023 applied Circuitscape 5 to a graph of 172 core areas and 424 links in Greater Geneva, mapping the major corridors and confirming the role of specific wildlife overpasses.

```{r}
#| label: fig-lcp-circuit-schematic
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Schematic contrast between least-cost path and circuit theory on the same landscape, with a source (S), a target (T), and a barrier that has a single gap. Left: least-cost path analysis returns only the single cheapest route. Right: circuit theory spreads movement across all parallel routes at once, and where the landscape forces flow through the narrow gap a pinch point forms (arrow), which least-cost analysis cannot reveal."

op <- par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1))

draw_land <- function(title) {
  plot(0, 0, type = "n", xlim = c(0, 10), ylim = c(0, 10), axes = FALSE,
       xlab = "", ylab = "", main = title)
  rect(0, 0, 10, 10, col = "#eef3ee", border = NA)
  rect(4, 0, 5.2, 4.2, col = "#6c757d", border = NA)
  rect(4, 5.8, 5.2, 10, col = "#6c757d", border = NA)
  points(1.5, 5, pch = 19, cex = 2.4, col = "#2d6a4f"); text(1.5, 5, "S", col = "white", cex = 0.8)
  points(8.5, 5, pch = 19, cex = 2.4, col = "#2d6a4f"); text(8.5, 5, "T", col = "white", cex = 0.8)
}

draw_land("Least-cost path")
lines(c(1.5, 3.5, 4.6, 5, 6.5, 8.5), c(5, 5.1, 5, 5, 4.9, 5), col = "#1d3557", lwd = 3)
text(4.6, 6.9, "barrier", col = "#495057", cex = 0.7)

draw_land("Circuit theory")
set.seed(4)
for (k in 1:16)
  lines(c(1.5, 3.5, 4.6, 5, 6.5, 8.5),
        c(5, 5 + rnorm(1, 0, 0.6), 4.95 + rnorm(1, 0, 0.12),
          5 + rnorm(1, 0, 0.1), 5 + rnorm(1, 0, 0.6), 5),
        col = rgb(0.8, 0.2, 0.1, 0.16), lwd = 2)
arrows(4.6, 7.6, 4.6, 5.4, length = 0.08, col = "#9b2226", lwd = 1.6)
text(4.6, 7.9, "pinch point", col = "#9b2226", cex = 0.7)
par(op)
```

### Infection-Risk Kernels and Cost-Distance Dispersal

An infection-risk kernel converts a cost-distance analysis into the probability that a dispersing infected animal reaches a given location before dying. Dispersal is described by an exponential kernel $P(d) = \exp(-d/D)$, where $d$ is a distance measure and $D$ is a characteristic dispersal scale set from telemetry. The single parameter $D$ has a direct probabilistic interpretation and can be read from the literature. For wild boar, the relevant scale during the infectious period is the mean natal dispersal distance rather than the home-range diameter, because infected animals move normally until the final clinical phase. Telemetry and mark-recapture studies across European populations report mean dispersal of approximately 4 kilometres, with most events within 10 kilometres of the birth site [@truve2003; @keuling2010; @prevot2013; @morelle2015]. Critically, $d$ is taken as Dijkstra cost-distance through the resistance surface rather than straight-line distance, so the kernel assigns near-zero risk to cells that are geographically close but separated by a motorway or large water body. The EFSA operational framework gives a validation benchmark [@efsa2018]: the 3-kilometre protection zone corresponds to a risk of about 0.50 under the kernel with $D = 4$ km, and the 10-kilometre surveillance boundary to about 0.08, which lets the displayed risk classes be anchored to recognised thresholds.

```{r}
#| label: fig-kernel-schematic
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.5
#| fig-cap: "Schematic of the cost-distance infection-risk kernel. Risk decays exponentially from the outbreak origin. Because distance is measured through the resistance surface rather than in a straight line, a barrier casts a low-risk shadow: location B, behind the barrier, carries far lower risk than location A at the same straight-line distance in open terrain. Dashed and dotted circles mark the EFSA protection (about 3 km) and surveillance (about 10 km) thresholds."

par(mar = c(1, 1, 2, 1))
plot(0, 0, type = "n", xlim = c(-10, 10), ylim = c(-10, 10), axes = FALSE,
     xlab = "", ylab = "", asp = 1, main = "Cost-distance infection-risk kernel")
rr  <- seq(10, 0.5, by = -0.5)
pal <- colorRampPalette(c("#fff7ec", "#fdbb84", "#d7301f"))(length(rr))
for (i in seq_along(rr)) symbols(0, 0, circles = rr[i], inches = FALSE, add = TRUE, bg = pal[i], fg = NA)
polygon(c(3.3, 11, 11, 3.3), c(-9, -9, 3, 3), col = "#fff7ec", border = NA)  # cost-distance shadow
rect(2.5, -9, 3.3, 3, col = "#6c757d", border = NA); text(2.9, -9.7, "barrier", cex = 0.65)
symbols(0, 0, circles = 3,  inches = FALSE, add = TRUE, fg = "black", lty = 2, lwd = 1.2)
symbols(0, 0, circles = 10, inches = FALSE, add = TRUE, fg = "black", lty = 3, lwd = 1.2)
text(-2.7, 2.4, "3 km", cex = 0.6); text(-6.6, 7.1, "10 km", cex = 0.6)
points(0, 0, pch = 19, cex = 1.1); text(0, -1, "origin", cex = 0.65)
points(-4, 0, pch = 17, col = "#1d3557", cex = 1.3); text(-4, 0.9, "A", cex = 0.85)
points(6, -4, pch = 17, col = "#1d3557", cex = 1.3); text(6.4, -4.8, "B", cex = 0.85)
```

### Biased Correlated Random Walk

A biased correlated random walk (BCRW) models movement as discrete steps shaped by two forces: directional memory, which correlates successive steps, and an environmental bias toward low-cost terrain [@turchin1998; @codling2008]. Directional persistence is described by the von Mises distribution, where the probability of a turn $\theta$ is proportional to $\exp(\kappa \cos\theta)$ and the concentration $\kappa$ controls how strongly the walker continues in its current direction. Weighting each candidate step by the inverse of destination resistance and step length adds the cost bias. As the number of walks grows, the BCRW visit density converges toward the circuit-theory current map under the Randomized Shortest Paths equivalence [@saerens2009], so the two methods estimate the same quantity by different means. In practice this limit is reached only with very many walks, which makes the BCRW a useful complementary diagnostic rather than a replacement for circuit theory.

```{r}
#| label: fig-bcrw-schematic
#| echo: false
#| fig-width: 6.5
#| fig-height: 5
#| fig-cap: "Schematic of a biased correlated random walk. Several stochastic trajectories start from the same source. Directional persistence keeps successive steps roughly aligned, while a bias toward low-cost terrain steers the walks along the low-resistance corridor (green band). Accumulated over many walks, the visit density approximates the circuit-theory current map."

par(mar = c(1, 1, 2, 1))
plot(0, 0, type = "n", xlim = c(0, 12), ylim = c(0, 10), axes = FALSE,
     xlab = "", ylab = "", main = "Biased correlated random walk")
rect(0, 0, 12, 10, col = "#ededed", border = NA)
rect(0, 4, 12, 6.2, col = "#cdebd0", border = NA)
text(10.4, 5.1, "low-cost\ncorridor", cex = 0.65, col = "#2d6a4f")
set.seed(7)
sx <- 1; sy <- 5
for (w in 1:8) {
  x <- sx; y <- sy; head <- 0
  for (s in 1:25) {
    head <- head + rnorm(1, 0, 0.5) + (5 - y) * 0.05
    nx <- x + 0.45 * cos(head); ny <- y + 0.45 * sin(head)
    segments(x, y, nx, ny, col = rgb(0.1, 0.3, 0.55, 0.6), lwd = 1.3)
    x <- nx; y <- ny
  }
}
points(sx, sy, pch = 19, cex = 1.3); text(sx, sy - 0.7, "source", cex = 0.7)
```

### Individual-Based Movement Models

Individual-based movement models take a finer-grained approach by reconstructing movement one step at a time and asking which landscape conditions at the end of each step the animal preferentially selects [@avgar2016]. The integrated step-selection function (iSSF) introduced above is the most widely used method in this class, and the same machinery that calibrates the resistance surface can also drive a forward simulation of individual trajectories. Rather than solving for a global flow as circuit theory does, an iSSF-based individual-based movement model (IBMM) releases simulated agents into the landscape and lets each one move according to the fitted step-length and turning-angle kernels, selecting successive cells in proportion to their estimated habitat suitability. Because the suitability surface is recovered directly from the iSSF coefficients, the simulated paths reflect the same empirically estimated preferences that define the resistance surface, which makes the IBMM a behaviourally explicit complement to the aggregate corridor methods. Its particular value for disease work is that each agent can be given a finite lifespan drawn from an infectious-period distribution, so the accumulated visit pattern estimates how far an infected animal is likely to spread contamination before dying rather than only where movement is possible [@avgar2016].

```{r}
#| label: fig-ibmm-schematic
#| echo: false
#| fig-width: 6.5
#| fig-height: 5
#| fig-cap: "Schematic of the iSSF-based individual movement model. Infected agents are released from the outbreak origin and move by the fitted step-selection rules. Each agent runs for a finite infectious period drawn from a distribution, so trajectories end (crosses) at different distances. The cells visited before death accumulate into a contamination-probability surface."

par(mar = c(1, 1, 2, 1))
plot(0, 0, type = "n", xlim = c(0, 12), ylim = c(0, 10), axes = FALSE,
     xlab = "", ylab = "", main = "iSSF-based individual movement model")
set.seed(11)
sx <- 1.5; sy <- 5
cols <- c("#1d3557", "#2a9d8f", "#e76f51", "#6a4c93", "#457b9d", "#bc6c25")
for (w in 1:6) {
  x <- sx; y <- sy; head <- runif(1, -0.4, 0.4)
  life <- sample(8:30, 1)
  for (s in 1:life) {
    head <- head + rnorm(1, 0, 0.45) + (5 - y) * 0.02
    nx <- x + 0.5 * cos(head); ny <- y + 0.5 * sin(head)
    segments(x, y, nx, ny, col = cols[w], lwd = 1.4)
    x <- nx; y <- ny
  }
  points(x, y, pch = 4, cex = 1.1, lwd = 2, col = cols[w])
}
points(sx, sy, pch = 19, cex = 1.5)
text(sx, sy - 0.8, "outbreak origin", cex = 0.7)
text(11.5, 9.4, "x = agent dies", cex = 0.65, adj = 1)
par(mar = c(5, 4, 4, 2) + 0.1)
```

## Wild Boar Ecology Relevant to Connectivity

The wild boar (*Sus scrofa*) is a native Eurasian ungulate whose European populations have expanded strongly in recent decades, driven by milder winters, the spread of maize and other high-energy crops, reduced hunting pressure, and supplementary winter feeding [@guberti2019]. Wild boar are habitat generalists that use woodland, especially deciduous and mixed forest, as daytime resting cover and forage at night in agricultural fields, showing marked behavioural flexibility in fragmented and peri-urban landscapes [@morelle2015]. This forest dependence for cover and avoidance of human infrastructure is what a resistance surface must capture. Home ranges are usually well below 50 square kilometres and movement contracts where food is stable and abundant [@guberti2019; @morelle2015]. The epidemiologically important exception is natal dispersal by young males, which can cover much greater distances and, rarely, exceed 100 kilometres [@guberti2019], because a dispersing animal in the incubation phase can carry infection into previously unaffected areas.

## African Swine Fever and the Connectivity Trade-off

ASF is caused by a large DNA virus of the family *Asfarviridae* that in Europe affects only domestic pigs and wild boar [@guberti2019]. The genotype II strain of the current epidemic has a case fatality rate of 90 to 95 percent, most infected animals die within four to nine days of showing signs, and no licensed vaccine exists, so control depends on surveillance, movement restriction, and population management. The virus is exceptionally stable, remaining infectious in blood for up to 15 weeks and persisting in carcasses for months, so infected carcasses act as long-lived reservoirs that decouple persistence from host density [@guberti2019]. Transmission runs through direct contact, which depends on host density, and through contact with carcasses and fomites, which sustains the cycle even as the live population declines. Since its introduction to the Caucasus in 2007, ASF has advanced westward at roughly 1 to 3 kilometres per month through wild boar populations, with human-induced outbreaks that form new clusters ahead of the front, leaving currently disease-free countries including Switzerland at near-term risk [@guberti2019].

The more connected and continuous the wild boar population, the faster the epidemic spread advances, while natural or artificial barriers can slow it. This makes connectivity a genuine management variable: authorities can in principle reduce spread by reinforcing barriers, installing targeted fencing in high-connectivity areas, or lowering density in specific corridors. Doing so effectively requires knowing where the corridors and pinch points are and predicting how an intervention would change them. @hess1994 provided formal support for this and demonstrated that conservation corridors generally aid metapopulation persistence but, under conditions of high direct-contact transmission and high disease-induced mortality that closely match ASF, can substantially raise the risk of disease-driven extinction. The same corridors that benefit red deer can therefore accelerate ASF in wild boar. Existing models, including those of @fischer2024 and @urbina2023, are rigorous but static and cannot answer what-if questions about a proposed fence or overpass. Building a tool that combines their methodological rigour with interactive, user-defined scenarios for wild boar in the Canton of Zürich is the central task of this thesis.
