# Results

Two primary deliverables were created during this project: a 25-metre resolution wild boar resistance surface covering Canton Aargau (used for model calibration) and Canton Zürich (produced by parameter transfer), and the *Wildboar Connectivity (ASF)* QGIS plugin that uses this resistance surface as its analytical input. The following sections present the results in the order of their production: from telemetry data quality and step classification through spatial covariate construction, iSSF model fitting, and resistance surface generation, concluding with a functional demonstration of the interactive plugin.

```{r}
#| label: setup
#| include: false

# ── Project paths ──────────────────────────────────────────────────────────
project_root <- "C:/ZHAW/6.Semester/BA/BA-wild-boar-connectivity-modeling"
data_temp    <- file.path(project_root, "data/temp")
data_proc    <- file.path(project_root, "data/processed")
data_raw     <- file.path(project_root, "data/raw")

# ── Packages ───────────────────────────────────────────────────────────────
options(repos = c(CRAN = "https://cloud.r-project.org"))
pkgs <- c("terra", "ggplot2", "dplyr", "knitr", "kableExtra",
          "patchwork", "scales", "leaflet", "raster")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}

# ── Global knitr options ───────────────────────────────────────────────────
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi     = 150
)
terraOptions(progress = 0)  # suppress progress bars

# ── Helper functions ───────────────────────────────────────────────────────

# Load a raster and aggregate for display
disp <- function(path, fact = 5, fun = "mean") {
  r <- rast(path)
  if (fact > 1) aggregate(r, fact = fact, fun = fun, na.rm = TRUE) else r
}

# Load a binary raster, aggregate with modal, set 0 to NA for display
disp_binary <- function(path, fact = 5) {
  r <- rast(path)
  r_agg <- if (fact > 1) aggregate(r, fact = fact, fun = "modal", na.rm = TRUE) else r
  r_agg[r_agg[] == 0] <- NA
  r_agg
}

# Compute summary statistics for a terra SpatRaster
raster_stats <- function(r, label = "") {
  v <- values(r, mat = FALSE, na.rm = TRUE)
  if (length(v) == 0) return(NULL)
  total <- prod(dim(r)[1:2])
  data.frame(
    Canton        = label,
    `Min`         = round(min(v), 3),
    `Mean`        = round(mean(v), 3),
    `Median`      = round(median(v), 3),
    `SD`          = round(sd(v), 3),
    `P10`         = round(quantile(v, 0.10), 3),
    `P90`         = round(quantile(v, 0.90), 3),
    `Max`         = round(max(v), 3),
    `Valid cells` = format(length(v), big.mark = ","),
    `NoData (%)`  = round(100 * (total - length(v)) / total, 1),
    check.names   = FALSE
  )
}

# Load the full 14-band Aargau covariate stack once
stk <- rast(file.path(data_temp, "covariate_stack.tif"))
names(stk) <- c("forest", "forest_density", "forest_density_g",
                "dist_forest_edge", "slope", "tri",
                "dist_flowing", "dist_lakes", "dist_water",
                "dist_fenced_road", "dist_main_road",
                "dist_secondary_road", "dist_settlement", "dist_railway")
```

## Resistance Surface

In the following section the results of creating the resistance surface are shown and explained in detail.

### Telemetry Quality and Step Classification

Raw GPS telemetry data from 16 wild boar tracked in Canton Aargau between 2014 and 2017 were processed prior to trajectory analysis. Individuals with a tracking duration of less than 30 days or fewer than 100 total fixes were excluded from the dataset (n = 2). For the remaining 14 individuals (yielding 161,275 raw fixes), the trajectories were resampled to standardise the sampling frequency. A nominal 15-minute fix interval with a 5-minute tolerance was applied. Fixes that fell outside this 10 to 20-minute window, or redundant fixes within the same window, were dropped. This regularisation resulted in a 14.3% data loss (23,015 excluded fixes). Additionally, three biologically unrealistic step lengths (\>3000m) were also dropped. The remaining 138,260 standardised fixes were subsequently converted into steps, yielding a final dataset of 133,109 continuous steps for analysis.

The three-rule hierarchical classifier partitioned these steps into 7,824 (5.9%) in-matrix transit steps and 125,285 (94.1%) in-patch resting steps. Among in-matrix steps, Rule 1 (step length greater than 250 metres) captured 4,759 steps (60.8% of all in-matrix steps), Rule 2 (step length greater than 150 metres and absolute turning angle below π/4 radians) contributed 2,748 steps (35.1%), and Rule 3 (three or more consecutive directional steps each exceeding 100 metres) added 317 steps (4.1%). The in-matrix dataset served as the sole analytical sample for iSSF fitting. In-patch steps were excluded from the model because they describe resting and foraging behaviour within habitat patches rather than landscape-level transit between patches, and combining the two behavioural states would confound the resistance surface.

In-matrix step lengths ranged from 100 metres to 5,424 metres with a mean of 299 metres, a median of 273 metres, and a standard deviation of 163 metres. The distribution is right-skewed, consistent with the gamma-distributed step-length kernel assumed by the iSSF movement model. Turning angles were approximately symmetric around zero with a standard deviation of approximately 59 degrees, indicating moderate directional persistence during transit. Which shows that the step classification worked.

```{r}
#| label: tbl-classification
#| tbl-cap: "Step classification counts for the in-matrix telemetry dataset. Rules are applied hierarchically in the order shown. A step is classified as in-matrix if it meets any rule that has not been met by a prior rule."

rule_df <- data.frame(
  Rule = c("Rule 1", "Rule 2", "Rule 3", "Total in-matrix", "In-patch"),
  Criterion = c(
    "Step length > 250 m",
    "Step length > 150 m AND turning angle < π/4",
    "≥ 3 consecutive steps: length > 100 m AND turning angle < π/4",
    "—",
    "All remaining steps (conservative default)"
  ),
  `Steps (n)` = c(4752, 2741, 317, 7810, 124578),
  `Share of in-matrix (%)` = c(60.8, 35.1, 4.1, 100.0, NA),
  check.names = FALSE
)
rule_df[is.na(rule_df)] <- "—"

kable(rule_df, align = c("l", "l", "r", "r"), booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position", "striped"), full_width = FALSE) |>
  row_spec(4, bold = TRUE) |>
  row_spec(5, italic = TRUE)
```

```{r}
#| label: fig-step-distributions
#| fig-cap: "Movement statistics of wild boar steps used for iSSF fitting. **Top Row (In-matrix):** Distribution of step lengths (left) and turning angles (right) during transit. **Bottom Row (In-patch):** Distribution of step lengths (left) and turning angles (right) during resident/foraging behavior. Vertical lines on the step length plots mark the classification thresholds at 100 m, 150 m, and 250 m used by Rules 3, 2, and 1. Dashed vertical lines on the turning angle plots mark -π/4 and π/4 radians. Symmetric, unimodal distributions near zero in turning angles indicate directional persistence."
#| fig-width: 11
#| fig-height: 8.4

# Load both datasets
df_in    <- read.csv(file.path(data_proc, "WildBoar_InMatrix_Steps.csv"))
df_patch <- read.csv(file.path(data_proc, "WildBoar_InPatch_Steps.csv"))

# === TOP ROW: IN-MATRIX PLOTS ===

p_sl_in <- ggplot(df_in, aes(x = sl_)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#2d6a4f", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = c(100, 150, 250),
             linetype = c("dotted", "dashed", "solid"),
             colour   = c("#e9c46a", "#f4a261", "#e76f51"),
             linewidth = 0.8) +
  scale_x_continuous(labels = scales::comma, limits = c(0, 1500)) +
  labs(x = "Step length (m)", y = "Density",
       title = "In-matrix step lengths") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(size = 11))

p_ta_in <- ggplot(df_in, aes(x = ta_)) +
  geom_histogram(bins = 60, fill = "#457b9d", colour = "white", alpha = 0.85) +
  scale_x_continuous(
    breaks = c(-pi, -pi/2, 0, pi/2, pi),
    labels = c("-π", "-π/2", "0", "π/2", "π")
  ) +
  geom_vline(xintercept = c(-pi/4, pi/4),
             linetype = "dashed", colour = "#e76f51", linewidth = 0.7) +
  labs(x = "Turning angle (radians)", y = "Count",
       title = "In-matrix turning angles") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(size = 11))

# === BOTTOM ROW: IN-PATCH PLOTS ===

p_sl_patch <- ggplot(df_patch, aes(x = sl_)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#2d6a4f", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = c(100, 150, 250),
             linetype = c("dotted", "dashed", "solid"),
             colour   = c("#e9c46a", "#f4a261", "#e76f51"),
             linewidth = 0.8) +
  scale_x_continuous(labels = scales::comma, limits = c(0, 1500)) +
  labs(x = "Step length (m)", y = "Density",
       title = "In-patch step lengths") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(size = 11))

p_ta_patch <- ggplot(df_patch, aes(x = ta_)) +
  geom_histogram(bins = 60, fill = "#457b9d", colour = "white", alpha = 0.85) +
  scale_x_continuous(
    breaks = c(-pi, -pi/2, 0, pi/2, pi),
    labels = c("-π", "-π/2", "0", "π/2", "π")
  ) +
  geom_vline(xintercept = c(-pi/4, pi/4),
             linetype = "dashed", colour = "#e76f51", linewidth = 0.7) +
  labs(x = "Turning angle (radians)", y = "Count",
       title = "In-patch turning angles") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(size = 11))

# === COMBINE PLOTS ===
(p_sl_in + p_ta_in) / (p_sl_patch + p_ta_patch)
```

### Intermediate Spatial Covariate Layers

Fourteen spatial covariates were assembled on the 25-metre Aargau grid before VIF testing. These are presented below in functional groups. Layers included in the final iSSF model are labelled accordingly. Excluded layers are shown to document the full intermediate output of the pipeline and to provide spatial context for understanding the final resistance surface structure. For all distance rasters, values are in metres measured as Euclidean distance to the nearest cell of the target feature class. For all density and suitability rasters, values range from 0 to 1. Binary barrier and passage rasters contain value 1 where the feature is present and are stored as NoData elsewhere.

#### Forest Covariates

The forest cover layer classified 41.0% of valid cells as forest, reflecting the mosaic of wooded slopes and agricultural lowlands characteristic of the Aargau study area. Forest density at the 50-metre scale is a continuous proportion expressing local canopy cover intensity, and at a mean of 0.41 it closely mirrors the binary cover fraction. The 1,000-metre forest density gradient captures the same quantity smoothed over a broader landscape window, producing a spatially coarser surface that records the transition from the forested Jura foothills to the open Mittelland plain. Both density layers were included in the final iSSF model. The binary forest cover and the forest-edge distance layer were excluded during VIF testing.

```{r}
#| label: fig-forest-covariates
#| fig-cap: "Forest-related covariate layers for the Aargau study area at aggregated display resolution (125 m). **Top Left:** Binary forest cover. **Top Right:** Distance to forest edge (m). **Bottom Left:** Forest density at the 50-metre scale. **Bottom Right:** Forest density at the 1,000-metre scale. The two density layers (bottom row) were retained in the final iSSF model, whereas binary forest cover and distance to edge (top row) were excluded during VIF screening due to high collinearity."
#| fig-width: 10
#| fig-height: 8.5

green_pal <- hcl.colors(100, "Greens")

par(mfrow = c(2, 2), mar = c(1.5, 1.5, 12, 1.5))

# 1. Forest binary (Top Left) — non-forest = lightest green, forest = darkest
r_for <- aggregate(stk[["forest"]], 5, "modal", na.rm = TRUE)
plot(r_for, col = c(green_pal[1], green_pal[100]),
     main = "Forest Cover (binary)\n[excl. by VIF]",
     legend = FALSE, axes = FALSE)
legend("bottomleft", fill = c(green_pal[1], green_pal[100]),
       legend = c("Non-forest", "Forest"), cex = 0.75, bty = "n")

# 2. Distance to forest edge (Top Right) — low distance = darkest green (rev = TRUE)
r_edge <- aggregate(stk[["dist_forest_edge"]], 5, "mean", na.rm = TRUE)
plot(r_edge, col = hcl.colors(100, "Greens", rev = FALSE),
     main = "Distance to Forest Edge (m)\n[excl. by VIF]",
     axes = FALSE, plg = list(cex = 0.75))

# 3. Forest density 50m (Bottom Left) — high density = darkest green
r_fd <- aggregate(stk[["forest_density"]], 5, "mean", na.rm = TRUE)
plot(r_fd, col = green_pal,
     main = "Forest Density — 50 m\n[included in iSSF]",
     axes = FALSE, plg = list(cex = 0.75))

# 4. Forest density gradient 1000m (Bottom Right) — high density = darkest green
r_fdg <- aggregate(stk[["forest_density_g"]], 5, "mean", na.rm = TRUE)
plot(r_fdg, col = green_pal,
     main = "Forest Density Gradient — 1000 m\n[included in iSSF]",
     axes = FALSE, plg = list(cex = 0.75))
```

#### Terrain Covariates

Slope and terrain ruggedness index were computed from the DHM25 digital elevation model at 25-metre resolution. Slope in the study area ranged from 0 to 69 degrees with a mean of 8.1 degrees, reflecting the gentle topography of the Aargau lowlands with steeper values confined to the Jura slopes. TRI ranged from 0 to 54.5 with a mean of 2.9, indicating overall low topographic roughness. Both layers were excluded from the final iSSF model because of extreme collinearity (VIF exceeding 80 in both cases).

```{r}
#| label: fig-terrain-covariates
#| fig-cap: "Terrain-related covariate layers for the Aargau study area at aggregated display resolution (125 m). **Left:** Slope (degrees), indicating the steepness of the terrain. **Right:** Terrain Ruggedness Index (TRI), measuring local elevation variability. Both layers were derived from a digital elevation model and used as candidate covariates in the iSSF model."
#| fig-width: 10
#| fig-height: 4.5

terrain_pal <- hcl.colors(100, "Lajolla")

par(mfrow = c(1, 2), mar = c(1.5, 1.5, 4.5, 5))

# 1. Slope (Left)
r_slope <- aggregate(stk[["slope"]], 5, "mean", na.rm = TRUE)
plot(r_slope, col = terrain_pal,
     main = "Slope (degrees)",
     axes = FALSE, plg = list(cex = 0.75))

# 2. TRI (Right)
r_tri <- aggregate(stk[["tri"]], 5, "mean", na.rm = TRUE)
plot(r_tri, col = terrain_pal,
     main = "Terrain Ruggedness Index (TRI)",
     axes = FALSE, plg = list(cex = 0.75))
```

#### Water Covariates

Distance to water was the final covariate used in the iSSF model. It was derived by merging the distance-to-flowing-water and distance-to-lake layers into a single combined distance-to-any-water surface by taking the cell-wise minimum of the two component rasters. Distance to flowing water alone ranged from 0 to 3,466 metres (mean 242 m), reflecting the dense stream network of the Aargau lowlands. Distance to lakes was considerably larger on average (mean 978 m, maximum 7,164 m) because standing water bodies are spatially sparse in the study area. The combined distance-to-water surface (mean 225 m) therefore largely reflects proximity to streams. The component layers were excluded after merging to avoid double-counting the water feature class in the VIF testing.

```{r}
#| label: fig-water-covariates
#| fig-cap: "Water proximity covariates. **Left:** Distance to flowing water (m). **Centre:** Distance to lakes (m). **Right:** Combined distance to any water body (m), used in the final iSSF model. The combined layer was derived as the cell-wise minimum of the two component layers."
#| fig-width: 12
#| fig-height: 4.2

water_pal <- hcl.colors(100, "Blues", rev = FALSE)

par(mfrow = c(1, 3), mar = c(1.5, 1.5, 4.5, 1.5))

r_dfl <- aggregate(stk[["dist_flowing"]], 5, "mean", na.rm = TRUE)
plot(r_dfl, col = water_pal,
     main = "Distance to Flowing Water (m)\n[excluded after merging]",
     axes = FALSE, plg = list(cex = 0.75))

r_dlk <- aggregate(stk[["dist_lakes"]], 5, "mean", na.rm = TRUE)
plot(r_dlk, col = water_pal,
     main = "Distance to Lakes (m)\n[excluded after merging]",
     axes = FALSE, plg = list(cex = 0.75))

r_dw <- aggregate(stk[["dist_water"]], 5, "mean", na.rm = TRUE)
plot(r_dw, col = water_pal,
     main = "Distance to Water — Combined (m)\n[included in iSSF]",
     axes = FALSE, plg = list(cex = 0.75))
```

#### Infrastructure Distance Covariates

The four infrastructure distance layers that entered the final iSSF model show strongly different spatial structures reflecting the hierarchy of the transport network. Distance to main road ranged from 0 to 7,092 metres with a mean of 747 metres, documenting that the Aargau study area is relatively well served by the cantonal and national road network. Distance to secondary road had a much lower mean of 160 metres and a maximum of only 3,250 metres, indicating that secondary roads permeate the agricultural lowlands at close intervals. Distance to railway had the largest mean (1,756 metres) and range (0 to 7,243 metres) of all infrastructure covariates, reflecting the coarser spacing of the rail network compared with roads. Distance to settlement ranged from 0 to 6,976 metres (mean 206 metres), with low values concentrated in the densely settled Aare valley floor.

```{r}
#| label: fig-infrastructure-covariates
#| fig-cap: "Infrastructure distance covariates included in the final iSSF model. All distances are in metres. Higher values (darker red) indicate greater separation from the respective infrastructure class, which the positive iSSF coefficients show to be strongly preferred by wild boar during transit."
#| fig-width: 12
#| fig-height: 8.5

infra_pal <- hcl.colors(100, "Reds")

par(mfrow = c(2, 2), mar = c(1.5, 1.5, 12, 1.5))

r_dmr <- aggregate(stk[["dist_main_road"]], 5, "mean", na.rm = TRUE)
plot(r_dmr, col = infra_pal,
     main = "Distance to Main Road (m)\n[included in iSSF, β = 0.244]",
     axes = FALSE, plg = list(cex = 0.75))

r_dsr <- aggregate(stk[["dist_secondary_road"]], 5, "mean", na.rm = TRUE)
plot(r_dsr, col = infra_pal,
     main = "Distance to Secondary Road (m)\n[included in iSSF, β = 0.171]",
     axes = FALSE, plg = list(cex = 0.75))

r_dra <- aggregate(stk[["dist_railway"]], 5, "mean", na.rm = TRUE)
plot(r_dra, col = infra_pal,
     main = "Distance to Railway (m)\n[included in iSSF, β = 0.505]",
     axes = FALSE, plg = list(cex = 0.75))

r_dse <- aggregate(stk[["dist_settlement"]], 5, "mean", na.rm = TRUE)
plot(r_dse, col = infra_pal,
     main = "Distance to Settlement (m)\n[included in iSSF, β = 0.210]",
     axes = FALSE, plg = list(cex = 0.75))
```

The distance-to-fenced-road layer, while part of the initial covariate set, was excluded from the iSSF model before VIF testing because of a recognised trapping artifact. Fenced motorway infrastructure in Aargau is spatially sparse (mean distance 4,179 metres, maximum 16,710 metres), and the few in-matrix GPS steps recorded near fenced segments were disproportionately concentrated there relative to the available landscape. This creates a positive correlation between fenced-road proximity and step selection that reflects the limited geographic extent of the feature rather than genuine habitat preference, and the coefficient was not considered transferable to Zürich where fenced motorway density is substantially higher.

The distance-to-fenced-road layer is mapped in the appendix (@fig-fenced-road).

#### Absolute Barrier and Wildlife Passage Layers

Four binary layers defined the cells that received NoData (impassable) or minimum resistance (passable) treatment in the Keeley burn-in step, independently of the iSSF. Highway cells (motorways, motorway approaches, exit ramps, and car-train carriers) and cells within large water bodies exceeding 1 km² in area were set to NoData. Infrastructure layers were assigned the maximum resistance value. Wildlife passage locations were set to the minimum resistance value R = 1, ensuring that validated crossing structures function as zero-cost corridors in the connectivity model. Together these four layers account for the 10.8% of cells coded as NoData in the final Aargau resistance raster.

```{r}
#| label: fig-barrier-layers-aargau
#| fig-cap: "Binary barrier and passage layers for the Aargau study area. Coloured cells indicate feature presence, and grey cells are NoData or non-feature areas. **Left:** Highway cells (motorways and access roads) set to absolute barriers. **Centre:** Large water bodies (area > 1 km²) set to absolute barriers. **Right:** Wildlife passage locations set to minimum resistance (R = 1)."
#| fig-width: 12
#| fig-height: 4.2

par(mfrow = c(1, 3), mar = c(1.5, 1.5, 4.5, 1.5))

r_hw <- aggregate(rast(file.path(data_temp, "highway.tif")), 5, "modal", na.rm = TRUE)
r_hw[r_hw[] == 0] <- NA
plot(r_hw, col = "#e74c3c", legend = FALSE,
     main = "Highway Barriers\n[→ NoData in resistance]", axes = FALSE)

r_ll <- aggregate(rast(file.path(data_temp, "large_lakes.tif")), 5, "modal", na.rm = TRUE)
plot(r_ll, col = "#2980b9", legend = FALSE,
     main = "Large Lakes (> 1 km²)\n[→ NoData in resistance]", axes = FALSE)

r_pa <- aggregate(rast(file.path(data_temp, "passages.tif")), 5, "modal", na.rm = TRUE)
r_pa[r_pa[] == 0] <- NA
plot(r_pa, col = "white", legend = FALSE, background = "black",
     main = "Wildlife Passages\n[→ R_min = 1 in resistance]", axes = FALSE)
```

For the Zürich resistance surface, an additional binary layer encoded the locations of permanent ASP fences erected by the Canton Zürich veterinary office. All cells intersecting a permanent ASP fence were set to NoData in the Zürich resistance raster. This layer was computed from the official shapefile provided by the Veterinary Office and applied as a Zürich-specific burn-in step with no equivalent in the Aargau pipeline.

### iSSF Model Coefficients and Variable Selection

Before fitting the final conditional logistic regression, seven covariates were removed from the original set of 14 based on a VIF (variance inflation factor) test. Slope and terrain ruggedness index (TRI) were insignificant and had VIF values exceeding 80, driven by their near-collinearity. Binary forest cover and Distance to forest Edge were removed because they were insignificant when the forest density layers were added. Distance to flowing water and distance to lakes were merged into a single combined distance-to-water layer before covariate assembly, and the two original layers were then removed as the new creates variable was more significant. Distance to fenced road was excluded because its spatial trapping pattern: in-matrix steps were concentrated near the few fenced road segments present, inflating the coefficient and making it non-transferable to other landscapes. The seven remaining covariates formed the final iSSF model.

The conditional logistic regression model demonstrated a good fit (Concordance = 0.640, SE = 0.004) and provided a significant improvement over a null model (Likelihood ratio test: $\chi^2 = 2072$, df = 9, $p < 0.001$). As expected for the in-matrix transit state, the baseline movement parameters were highly significant, indicating strong directional persistence ($\beta_{cos\_ta} = 0.562$, $p < 0.001$) and a tendency for longer step lengths ($\beta_{log\_sl} = 0.155$, $p < 0.001$).Regarding environmental features, all seven retained spatial selection coefficients were positive and statistically significant at the $\alpha = 0.05$ level, indicating that transiting wild boar preferentially selected landscape cells offering higher forest density, stronger forest density gradients, and greater distance from all infrastructure categories and water. The strongest overall selection effect was for forest density at the 50-metre scale ($\beta = 0.516$, $z = 29.31$), reflecting the central and highly consistent role of dense forest cover as movement habitat. Avoidance of railway infrastructure was the second strongest predictor ($\beta = 0.462$), though this effect was estimated with substantially lower precision ($z = 4.00$) than the preference for forest, likely reflecting less frequent interaction with rail lines. Forest density gradient at the 1,000-metre scale was the third-ranked predictor ($\beta = 0.269$), capturing a preference for cells situated within a landscape matrix of increasing forest cover rather than isolated patches. Distance to main roads ($\beta = 0.260$), distance to settlements ($\beta = 0.210$), and distance to secondary roads ($\beta = 0.165$) followed sequentially, expressing a clear behavioral avoidance of the built environment and vehicular infrastructure. Distance to water was the weakest predictor ($\beta = 0.049$, $p = 0.038$), which remains consistent with wild boar avoiding water crossings during directed transit movement.

![Standardized selection coefficients and 95% confidence intervals for the final Integrated Step Selection Function (iSSF) modeling wild boar in-matrix transit behavior. Positive values indicate a selection preference for higher covariate values (i.e., selection for denser forest, and avoidance of infrastructure and water by maintaining greater distances). Forest density at the 50-metre scale exhibits the strongest and most precise selection effect. The wider confidence interval for distance to railway reflects the relative spatial scarcity of this infrastructure in the training area. All variables were z-standardized prior to model fitting to allow for the direct comparison of effect sizes.](images/issf_coefficients.png)

### Resistance Surface: Canton Aargau

```{r}
#| label: tbl-stats-aargau
#| tbl-cap: "Summary statistics for the Aargau habitat suitability raster and the derived resistance raster. Statistics are computed over all valid (non-NoData) cells. The theoretical range of the resistance raster under the Keeley transformation with C = 4 is [1.00, exp(4)] = [1.00, 54.598]."

r_suit_ag <- rast(file.path(data_proc, "Habitat_Suitability_Wildboar_Aargau.tif"))
r_res_ag  <- rast(file.path(data_proc, "Resistance_Keeley_Wildboar_Aargau_25m.tif"))

stats_ag <- rbind(
  raster_stats(r_suit_ag, "Suitability"),
  raster_stats(r_res_ag,  "Resistance")
)
stats_ag$Canton <- c("Aargau — Suitability", "Aargau — Resistance")

kable(stats_ag, align = c("l","r","r","r","r","r","r","r","r","r"),
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position", "striped"), full_width = FALSE)
```

The Aargau resistance raster covers 2,224 × 2,405 pixels at 25-metre resolution in the Swiss LV95 coordinate system (EPSG:2056), with a 2-kilometre buffer applied around the cantonal boundary to avoid edge effects. Resistance values span the full theoretical range of the Keeley transformation: from 1.00 (maximum suitability, minimum resistance) to 54.60, which equals exp(4) within numerical precision and corresponds to cells with effectively zero iSSF selection weight. The median resistance of 34.22 lies well above the midpoint of the valid range (approximately 27.8), indicating that the study area is predominantly high-resistance terrain from the perspective of a dispersing wild boar. The 90th percentile of resistance is at the maximum resistance of 54.6.

Spatially, low-resistance cells (R \< 10, corresponding to suitability \> 0.43) are concentrated in the large continuous forest blocks on the south-facing Jura slopes and in the riparian forests along the Aare and Reuss rivers **(Figure X)**. These areas provide the dense, well-structured forest with low infrastructure proximity that the iSSF identified as the dominant selection signal. High-resistance cells (R approaching 54.60) dominate the burn in infrastructure features and all areas within short distances of the road, rail, and settlement networks that dissect the Mittelland. The 10.8% of cells stored as NoData correspond exclusively to absolute barriers: fenced motorway corridors and large lakes.

```{r}
#| label: fig-aargau-maps
#| fig-cap: "Habitat suitability (left) and resistance (right) rasters for the Canton Aargau study area at 25-metre resolution (displayed at 100 m for rendering performance). Suitability ranges from 0 (minimum, red) to 1 (maximum, green) and represents the iSSF selection surface estimated from GPS telemetry. Resistance is the inverse Keeley transformation of suitability and ranges from 1 (minimum, green) to 54.60 (maximum, red). NoData cells (absolute barriers: motorways, large lakes, large settlements) are shown in dark grey."
#| fig-width: 12
#| fig-height: 5.5

# Aggregate to ~100m for rendering
r_suit_plot <- aggregate(r_suit_ag, 4, "mean", na.rm = TRUE)
r_res_plot  <- aggregate(r_res_ag,  4, "mean", na.rm = TRUE)

df_suit <- as.data.frame(r_suit_plot, xy = TRUE)
colnames(df_suit)[3] <- "value"
df_res  <- as.data.frame(r_res_plot,  xy = TRUE)
colnames(df_res)[3]  <- "value"

p_suit <- ggplot(df_suit, aes(x = x, y = y, fill = value)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_gradientn(
    colours  = hcl.colors(200, "RdYlGn"),
    limits   = c(0, 1),
    na.value = "#2c2c2c",
    name     = "Suitability"
  ) +
  coord_equal() +
  theme_void(base_size = 11) +
  theme(legend.position   = "right",
        plot.title        = element_text(size = 11, hjust = 0.5),
        legend.key.height = unit(1.5, "cm")) +
  labs(title = "Habitat Suitability — Canton Aargau")

p_res <- ggplot(df_res, aes(x = x, y = y, fill = value)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_gradientn(
    colours  = rev(hcl.colors(200, "RdYlGn")),
    limits   = c(1, 55),
    na.value = "#2c2c2c",
    name     = "Resistance"
  ) +
  coord_equal() +
  theme_void(base_size = 11) +
  theme(legend.position   = "right",
        plot.title        = element_text(size = 11, hjust = 0.5),
        legend.key.height = unit(1.5, "cm")) +
  labs(title = "Resistance Surface — Canton Aargau")

p_suit + p_res
```

```{r}
#| label: fig-aargau-histogram
#| fig-cap: "Frequency distribution of resistance values in the Aargau study area (valid cells only, n = 4,768,720). The pronounced right skew and spike at R = 54.60 reflect the dominance of high-resistance agricultural and peri-urban terrain. The 10.8% of impassable cells (NoData) are excluded from this distribution."
#| fig-width: 7
#| fig-height: 3.8

# Sample for histogram (avoid reading all 4.7M values into memory for plotting)
set.seed(42)
v_ag <- as.data.frame(spatSample(r_res_ag, size = 200000, method = "random", na.rm = TRUE))
col_name <- names(v_ag)[1]

ggplot(v_ag, aes(x = .data[[col_name]])) +
  geom_histogram(bins = 80, fill = "#457b9d", colour = "white", alpha = 0.85,
                 linewidth = 0.2) +
  geom_vline(xintercept = c(10, 27.8, 34.25),
             linetype = c("solid", "dashed", "dotted"),
             colour   = c("#2d6a4f", "grey40", "#e76f51"),
             linewidth = 0.7) +
  annotate("text", x = c(11.5, 29, 36.5), y = Inf,
           label = c("R = 10\n(low)", "midpoint\n(27.8)", "median\n(34.3)"),
           vjust = 1.4, hjust = 0, size = 3,
           colour = c("#2d6a4f", "grey40", "#c0422a")) +
  scale_x_continuous(limits = c(1, 56), breaks = seq(0, 55, 10)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Resistance (R)", y = "Count (sample of 200,000 cells)",
       title = "Resistance value distribution — Canton Aargau") +
  theme_bw(base_size = 11)
```

### Transfer to Canton Zürich

```{r}
#| label: tbl-stats-zurich
#| tbl-cap: "Summary statistics for the Zürich habitat suitability and resistance rasters. The Zürich parameters were derived by transferring the Aargau iSSF coefficients and scaling parameters (eta_min, eta_max) without refitting."

r_suit_zh <- rast(file.path(data_proc, "Habitat_Suitability_Wildboar_Zurich_25m.tif"))
r_res_zh  <- rast(file.path(data_proc, "Resistance_Keeley_Wildboar_Zurich_25m.tif"))

stats_zh <- rbind(
  raster_stats(r_suit_zh, "Zürich — Suitability"),
  raster_stats(r_res_zh,  "Zürich — Resistance")
)

kable(stats_zh, align = c("l","r","r","r","r","r","r","r","r","r"),
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position", "striped"), full_width = FALSE)
```

The Zürich resistance raster covers 2,538 × 2,066 pixels at the same 25-metre resolution and LV95 projection as the Aargau calibration product. Its value range of 1.00 to 54.60 is identical to Aargau by construction, because the Keeley transformation preserves theoretical bounds regardless of the underlying landscape. The median resistance of 24.1 and mean of 28.23 are both lower than in Aargau. This is a direct consequence of the Zürich-specific suitability normalisation, which stretches the full \[0, 1\] suitability scale across the Zürich habitat range so that the best Zürich forest cells receive the minimum resistance of 1.00.

Spatially, the most suitable, low-resistance cells are concentrated in the Sihlwald forest reserve south of the city and along the forested ridgelines of the Zürichsee hills **(Figure Y)**. The covariates are z-standardized with the Aargau training statistics to preserve the calibrated meaning of the iSSF coefficients, while the final normalization utilizes the local linear-predictor range to maximize within-Zürich contrast. The 15.5% NoData fraction is higher than in Aargau, comprising motorway barriers, large lake polygons, and the addition of permanent ASF prevention fences situated within the densely developed Zürich extent.

```{r}
#| label: fig-zurich-maps
#| fig-cap: "Habitat suitability (left) and resistance (right) for the Canton Zürich study area, with the suitability normalisation anchored to the Zürich linear-predictor range so that the full scale is used for within-Zürich variation. Colour scales match those used in the Aargau figure. NoData cells (motorways and large lakes) are shown in dark grey."
#| fig-width: 12
#| fig-height: 5.5

r_suit_zh_plot <- aggregate(r_suit_zh, 4, "mean", na.rm = TRUE)
r_res_zh_plot  <- aggregate(r_res_zh,  4, "mean", na.rm = TRUE)

df_suit_zh <- as.data.frame(r_suit_zh_plot, xy = TRUE)
colnames(df_suit_zh)[3] <- "value"
df_res_zh  <- as.data.frame(r_res_zh_plot,  xy = TRUE)
colnames(df_res_zh)[3]  <- "value"

p_suit_zh <- ggplot(df_suit_zh, aes(x = x, y = y, fill = value)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_gradientn(
    colours  = hcl.colors(200, "RdYlGn"),
    limits   = c(0, 1),
    na.value = "#2c2c2c",
    name     = "Suitability"
  ) +
  coord_equal() +
  theme_void(base_size = 11) +
  theme(legend.position   = "right",
        plot.title        = element_text(size = 11, hjust = 0.5),
        legend.key.height = unit(1.5, "cm")) +
  labs(title = "Habitat Suitability: Canton Zürich")

p_res_zh <- ggplot(df_res_zh, aes(x = x, y = y, fill = value)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_gradientn(
    colours  = rev(hcl.colors(200, "RdYlGn")),
    limits   = c(1, 55),
    na.value = "#2c2c2c",
    name     = "Resistance"
  ) +
  coord_equal() +
  theme_void(base_size = 11) +
  theme(legend.position   = "right",
        plot.title        = element_text(size = 11, hjust = 0.5),
        legend.key.height = unit(1.5, "cm")) +
  labs(title = "Resistance Surface: Canton Zürich")

p_suit_zh + p_res_zh
```

```{r}
#| label: fig-canton-comparison
#| fig-cap: "Resistance value distributions for Canton Aargau and Canton Zürich (random sample of 200,000 valid cells per canton). The Zürich distribution is shifted toward lower resistance values, visible in the lower median (23.83 vs 34.25), because the Zürich-specific normalisation uses the full resistance scale across the Zürich habitat range. Both cantons retain a substantial high-resistance fraction corresponding to the fragmented agricultural and built matrix."
#| fig-width: 8
#| fig-height: 4.2

set.seed(42)
v_zh <- spatSample(r_res_zh, size = 200000, method = "random", na.rm = TRUE)
v_ag_s <- spatSample(r_res_ag, size = 200000, method = "random", na.rm = TRUE)

df_comp <- rbind(
  data.frame(resistance = as.numeric(v_ag_s[[1]]), Canton = "Aargau (calibration)"),
  data.frame(resistance = as.numeric(v_zh[[1]]),   Canton = "Zürich (transfer)")
)

ggplot(df_comp, aes(x = resistance, fill = Canton, colour = Canton)) +
  geom_density(alpha = 0.45, linewidth = 0.8) +
  geom_vline(
    data = data.frame(
      Canton = c("Aargau (calibration)", "Zürich (transfer)"),
      med    = c(34.25, 23.83)
    ),
    aes(xintercept = med, colour = Canton),
    linetype = "dashed", linewidth = 0.8
  ) +
  scale_fill_manual(values  = c("Aargau (calibration)" = "#2d6a4f",
                                "Zürich (transfer)"   = "#457b9d")) +
  scale_colour_manual(values = c("Aargau (calibration)" = "#2d6a4f",
                                 "Zürich (transfer)"   = "#457b9d")) +
  scale_x_continuous(limits = c(1, 56), breaks = seq(0, 55, 10)) +
  labs(x = "Resistance (R)", y = "Density",
       fill = "Canton", colour = "Canton",
       title = "Resistance distributions: Aargau vs. Zürich") +
  annotate("text", x = c(36.5, 21.5), y = 0.0195,
           label = c("Aargau\nmedian\n34.3", "Zürich\nmedian\n23.8"),
           size = 3, colour = c("#1a4d30", "#2c5b7a")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top")
```

## QGIS Plugin

The *Wildboar Connectivity (ASF)* plugin was implemented as a QGIS 3.40 extension and tested against the Zürich resistance raster with the WILMA core habitat polygons as LCP destination targets. Functional testing confirmed that all five analytical modules produced spatially coherent outputs for a range of outbreak origin locations across the canton. Two scenarios are reported here: a baseline analysis at an unmodified outbreak location in a forested area, and a scenario comparison demonstrating the effect of drawing a virtual fence across a corridor identified by the pinchpoint analysis.

### Baseline Connectivity Analysis

For the baseline demonstration, the outbreak origin was placed in a forested grid cell north of the Winterthur city centre representative of a scenario in which the disease enters through a wild boar population in that region and disperses in different directions. All five outputs were enabled, with 500 BCRW realisations and 2,000 iSSF-IBMM agents. The analysis completed in under three minutes on a standard laptop.

The LCP corridor analysis connected the outbreak origin to the nearest reachable WILMA habitat clusters via the pre-computed Dijkstra cost grid. The graduated blue styling assigned light sky blue to distant corridors serving high cummulative resistance cost, vivid blue to intermediate connectors, and deep navy to the backbone segments shared by multiple routes and smaller cummulative resistance cost. The paths concentrated in the forested ridgelines and crossed the wildlife overpasses, consistent with the visual pattern of low-resistance cells in the Zürich resistance map.

The circuit-theory pinchpoint raster, computed using Circuitscape.jl via Julia subprocess, showed concentrated current density at the defined outbreak point and at several bottlenecks. These bottlenecks included, wildlife crossings, but also narrow forest corridors that are separated from one another by roads or railway tracks between habitat patches. These pinchpoints represent the locations where a fence intervention would produce the largest reduction in connectivity at lowest material cost.

The continuous infection-risk raster produced three discrete EFSA-aligned zones around the origin. The high-risk zone (risk above 0.50) extended approximately 2.5 to 3.5 kilometres in the least-resistant directions, consistent with its calibration to the 4-kilometre mean wild boar dispersal distance. The EFSA surveillance zone boundary (risk 0.15) was reached at cost-equivalent distances of approximately 8 to 10 kilometres through continuous forest and at shorter Euclidean distances across fragmented terrain.

The BCRW density and iSSF-IBMM contamination probability layers broadly corroborated the LCP and pinchpoint structure. The BCRW density surface showed the highest visitation counts along the same corridors identified by the Circuitscape analysis, confirming their role as preferred transit routes. The iSSF-IBMM log10-transformed contamination probability surface showed the steepest probability gradients at the forest-to-infrastructure transitions where resistance increases rapidly.

![Example outputs of the wildboar connectivity plugin for a simulated outbreak origin north of Winterthur. Top left: least-cost path corridors to the 100 nearest low-resistance habitat clusters, styled by cost-distance weight, with thick dark-blue segments indicating near-origin backbone corridors and thin light-blue segments indicating distant, high-cost connections. Top right: infection risk surface derived from an exponential decay function, classified into three zones following EFSA guidelines (high risk \> 0.50, medium risk 0.15–0.50, low risk ≤ 0.15). Bottom left: dispersal corridor probability raster from circuit theory, where bright cells indicate pinchpoints through which movement is concentrated. Bottom right: random-walk visit density from 500 biased correlated random walks, showing the stochastic spread envelope around the outbreak origin.](images/baseline.png)

### Scenario Comparison: Virtual Fence Intervention

To demonstrate the scenario comparison functionality, a virtual fence was drawn over a wildlife overpass and a new wildlife overpass was built in the north of winterthur crossing a highway. The fence was rasterized as a two-cell-wide NaN wall in the modified resistance array and the full analysis was rerun without changing the outbreak origin or any other parameter. Both runs used identical settings so that all differences in the output layers can be attributed entirely to the fence modification.

The pinchpoint raster showed a clear rerouting of current density away from the blocked corridor segment. The cumulative flow that previously concentrated at the motorway crossing was redistributed to the new overpass, building an alternative corridor. The LCP backbone similarly shifted away from the blocked crossing to the new generated one, increasing connectivity to the south. The infection-risk zone contracted asymmetrically on the side of the fence facing away from the outbreak origin, indicating reduced modelled reachability in the fenced direction.

![Effect of landscape modifications on dispersal corridors and pinchpoints for a simulated outbreak north of Winterthur. Left column shows the baseline scenario without interventions, right column shows the same outbreak after placement of a fence (red line) and a wildlife overpass (blue dot) on the resistance surface. Top row: least-cost path corridors to the 100 nearest habitat destinations, styled by cost-distance weight. The fence severs several southward corridor branches in the modified scenario, while the overpass maintains connectivity across an otherwise impassable barrier. Bottom row: circuit-theory pinchpoint rasters showing dispersal corridor probability. The fence visibly reduces current flow along the blocked routes, and the overpass creates a new high-current channel at its location, demonstrating that both barrier and crossing interventions are correctly propagated through the resistance grid before analysis.](images/lcp_circuit_fence.png)

![Effect of landscape modifications on infection risk and individual-based movement model (IBMM) outputs for a simulated outbreak north of Winterthur. Left column shows the baseline scenario, right column shows the modified scenario with a fence (red line) cutting across the southern corridor and a wildlife overpass (blue dot) placed on a barrier. Top row: infection risk surfaces classified by exponential decay into three zones following EFSA guidelines. The fence reduces the southward extent of the medium-risk zone, while the overpass opens a new low-risk finger of connectivity in that direction. Bottom row: iSSF-IBMM contamination probability surfaces (log\\textsubscript{10} scale, 2000 agents). The fence visibly suppresses agent penetration into the southern urban fringe, and the overpass enables a new dispersal channel through the previously impassable barrier, reflected in the shifted spread envelope in the modified scenario.](images/risk_ibmm_fence.png)

The scenario comparison illustrates the core purpose of the plugin: rather than producing a single static map, it enables iterative exploration of management alternatives within the same GIS session. A wildlife manager can test different fence alignments, observe how each one redirects the modelled corridor network, and compare the resulting infection-risk zones, all without leaving the QGIS environment or re-running any preprocessing steps.

------------------------------------------------------------------------

:::: {.content-visible when-format="html"}
## Interactive Raster Viewer

The following tabbed maps allow interactive exploration of every spatial layer produced by the resistance surface pipeline. All rasters are displayed over a swisstopo Swissimage basemap. Layers within each tab can be toggled on and off using the layer control panel (top-right corner of each map). Rasters are displayed at approximately 200-metre resolution to keep the HTML document size manageable.

```{r}
#| label: leaflet-setup
#| include: false

# Aggregation factor for leaflet display (~200m from 25m source)
LF <- 8

# Helper: aggregate raster, reproject to WGS84, convert to raster package object
to_leaflet <- function(r, fact = LF, fun = "mean") {
  r_sm  <- aggregate(r, fact = fact, fun = fun, na.rm = TRUE)
  r_wgs <- project(r_sm, "EPSG:4326")
  raster::raster(r_wgs)
}

# Helper: binary raster for leaflet (set 0 → NA before aggregating)
to_leaflet_bin <- function(path, fact = LF) {
  r <- rast(path)
  r[r[] == 0] <- NA
  r_sm  <- aggregate(r, fact = fact, fun = "modal", na.rm = TRUE)
  r_wgs <- project(r_sm, "EPSG:4326")
  raster::raster(r_wgs)
}

# Pre-load and prepare all layers
lf_forest      <- to_leaflet(stk[["forest"]])
lf_fd          <- to_leaflet(stk[["forest_density"]])
lf_fdg         <- to_leaflet(stk[["forest_density_g"]])
lf_dfe         <- to_leaflet(stk[["dist_forest_edge"]])
lf_slope       <- to_leaflet(stk[["slope"]])
lf_tri         <- to_leaflet(stk[["tri"]])
lf_dw          <- to_leaflet(stk[["dist_water"]])
lf_dfl         <- to_leaflet(stk[["dist_flowing"]])
lf_dlk         <- to_leaflet(stk[["dist_lakes"]])
lf_dmr         <- to_leaflet(stk[["dist_main_road"]])
lf_dsr         <- to_leaflet(stk[["dist_secondary_road"]])
lf_dra         <- to_leaflet(stk[["dist_railway"]])
lf_dse         <- to_leaflet(stk[["dist_settlement"]])
lf_dfr         <- to_leaflet(stk[["dist_fenced_road"]])
lf_hw          <- to_leaflet_bin(file.path(data_temp, "highway.tif"))
lf_ll          <- to_leaflet(rast(file.path(data_temp, "large_lakes.tif")))
lf_pa          <- to_leaflet_bin(file.path(data_temp, "passages.tif"))
lf_suit_ag     <- to_leaflet(r_suit_ag)
lf_res_ag      <- to_leaflet(r_res_ag)

# Zürich layers
lf_fd_zh       <- to_leaflet(rast(file.path(data_temp, "forest_density_zurich.tif")))
lf_fdg_zh      <- to_leaflet(rast(file.path(data_temp, "forest_density_g_zurich.tif")))
lf_dw_zh       <- to_leaflet(rast(file.path(data_temp, "dist_water_zurich.tif")))
lf_dmr_zh      <- to_leaflet(rast(file.path(data_temp, "dist_main_road_zurich.tif")))
lf_dsr_zh      <- to_leaflet(rast(file.path(data_temp, "dist_secondary_road_zurich.tif")))
lf_dra_zh      <- to_leaflet(rast(file.path(data_temp, "dist_railway_zurich.tif")))
lf_dse_zh      <- to_leaflet(rast(file.path(data_temp, "dist_settlement_zurich.tif")))
lf_hw_zh       <- to_leaflet_bin(file.path(data_temp, "highway_zurich.tif"))
lf_ll_zh       <- to_leaflet(rast(file.path(data_temp, "large_lakes_zurich.tif")))
lf_pa_zh       <- to_leaflet_bin(file.path(data_temp, "passages_zurich.tif"))
lf_suit_zh     <- to_leaflet(r_suit_zh)
lf_res_zh      <- to_leaflet(r_res_zh)

# Colour palettes
pal_green  <- colorNumeric(hcl.colors(100, "Greens"),     domain = c(0, 1),   na.color = "transparent")
pal_blues  <- colorNumeric(hcl.colors(100, "Blues"),      domain = NULL,       na.color = "transparent")
pal_terrain<- colorNumeric(hcl.colors(100, "YlOrBr"),     domain = NULL,       na.color = "transparent")
pal_suit   <- colorNumeric(hcl.colors(100, "RdYlGn"),     domain = c(0, 1),   na.color = "transparent")
pal_res    <- colorNumeric(rev(hcl.colors(100, "RdYlGn")),domain = c(1, 55),  na.color = "transparent")

# Base tile options
base_groups <- c("Swisstopo (light)", "OpenStreetMap", "Satellite")
```

::: panel-tabset
### Forest Covariates

```{r}
#| label: map-forest

leaflet(options = leafletOptions(minZoom = 8)) |>
  addProviderTiles("CartoDB.Positron",   group = "Swisstopo (light)") |>
  addProviderTiles("OpenStreetMap",      group = "OpenStreetMap") |>
  addProviderTiles("Esri.WorldImagery",  group = "Satellite") |>
  addRasterImage(lf_forest, colors = colorNumeric(c("#f5f5e0","#1a5c2e"), 0:1, na.color="transparent"),
                 opacity = 0.8, group = "Forest Cover (binary) — Aargau") |>
  addRasterImage(lf_fd,  colors = pal_green, opacity = 0.8,
                 group = "Forest Density 50 m — Aargau") |>
  addRasterImage(lf_fdg, colors = pal_green, opacity = 0.8,
                 group = "Forest Density Gradient 1000 m — Aargau") |>
  addRasterImage(lf_dfe, colors = pal_blues, opacity = 0.8,
                 group = "Dist Forest Edge — Aargau [excluded VIF]") |>
  addRasterImage(lf_fd_zh,  colors = pal_green, opacity = 0.8,
                 group = "Forest Density 50 m — Zürich") |>
  addRasterImage(lf_fdg_zh, colors = pal_green, opacity = 0.8,
                 group = "Forest Density Gradient 1000 m — Zürich") |>
  addLayersControl(
    baseGroups    = base_groups,
    overlayGroups = c("Forest Cover (binary) — Aargau",
                      "Forest Density 50 m — Aargau",
                      "Forest Density Gradient 1000 m — Aargau",
                      "Dist Forest Edge — Aargau [excluded VIF]",
                      "Forest Density 50 m — Zürich",
                      "Forest Density Gradient 1000 m — Zürich"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup(c("Forest Cover (binary) — Aargau",
              "Forest Density Gradient 1000 m — Aargau",
              "Dist Forest Edge — Aargau [excluded VIF]",
              "Forest Density 50 m — Zürich",
              "Forest Density Gradient 1000 m — Zürich")) |>
  setView(lng = 8.15, lat = 47.38, zoom = 10) |>
  addLegend(pal = pal_green, values = c(0, 1),
            title = "Forest density (0–1)", position = "bottomright")
```

### Terrain Covariates

```{r}
#| label: map-terrain

pal_sl  <- colorNumeric(hcl.colors(100, "YlOrBr"), domain = c(0, 70),   na.color = "transparent")
pal_tri <- colorNumeric(hcl.colors(100, "YlOrBr"), domain = c(0, 55),   na.color = "transparent")

leaflet(options = leafletOptions(minZoom = 8)) |>
  addProviderTiles("CartoDB.Positron",  group = "Swisstopo (light)") |>
  addProviderTiles("OpenStreetMap",     group = "OpenStreetMap") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addRasterImage(lf_slope, colors = pal_sl,  opacity = 0.8,
                 group = "Slope (degrees) — excluded VIF") |>
  addRasterImage(lf_tri,   colors = pal_tri, opacity = 0.8,
                 group = "TRI — excluded VIF") |>
  addLayersControl(
    baseGroups    = base_groups,
    overlayGroups = c("Slope (degrees) — excluded VIF",
                      "TRI — excluded VIF"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  setView(lng = 8.15, lat = 47.38, zoom = 10) |>
  addLegend(pal = pal_sl, values = c(0, 70),
            title = "Slope (°)", position = "bottomright")
```

### Water Covariates

```{r}
#| label: map-water

pal_dw  <- colorNumeric(hcl.colors(100, "Blues"), domain = c(0, 3500), na.color = "transparent")
pal_dlk <- colorNumeric(hcl.colors(100, "Blues"), domain = c(0, 7200), na.color = "transparent")

leaflet(options = leafletOptions(minZoom = 8)) |>
  addProviderTiles("CartoDB.Positron",  group = "Swisstopo (light)") |>
  addProviderTiles("OpenStreetMap",     group = "OpenStreetMap") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addRasterImage(lf_dw,   colors = pal_dw,  opacity = 0.8,
                 group = "Dist Water — combined — Aargau [in iSSF]") |>
  addRasterImage(lf_dfl,  colors = pal_dw,  opacity = 0.8,
                 group = "Dist Flowing Water — Aargau [excluded]") |>
  addRasterImage(lf_dlk,  colors = pal_dlk, opacity = 0.8,
                 group = "Dist Lakes — Aargau [excluded]") |>
  addRasterImage(lf_dw_zh, colors = pal_dw, opacity = 0.8,
                 group = "Dist Water — combined — Zürich [in iSSF]") |>
  addLayersControl(
    baseGroups    = base_groups,
    overlayGroups = c("Dist Water — combined — Aargau [in iSSF]",
                      "Dist Flowing Water — Aargau [excluded]",
                      "Dist Lakes — Aargau [excluded]",
                      "Dist Water — combined — Zürich [in iSSF]"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup(c("Dist Flowing Water — Aargau [excluded]",
              "Dist Lakes — Aargau [excluded]",
              "Dist Water — combined — Zürich [in iSSF]")) |>
  setView(lng = 8.15, lat = 47.38, zoom = 10) |>
  addLegend(pal = pal_dw, values = c(0, 3500),
            title = "Distance (m)", position = "bottomright")
```

### Infrastructure Covariates

```{r}
#| label: map-infrastructure

pal_mr  <- colorNumeric(hcl.colors(100, "Blues"), domain = c(0, 7100), na.color = "transparent")
pal_sr  <- colorNumeric(hcl.colors(100, "Blues"), domain = c(0, 3250), na.color = "transparent")
pal_rw  <- colorNumeric(hcl.colors(100, "Blues"), domain = c(0, 7250), na.color = "transparent")
pal_se  <- colorNumeric(hcl.colors(100, "Blues"), domain = c(0, 7000), na.color = "transparent")
pal_fr  <- colorNumeric(hcl.colors(100, "Blues"), domain = c(0,16710), na.color = "transparent")

leaflet(options = leafletOptions(minZoom = 8)) |>
  addProviderTiles("CartoDB.Positron",  group = "Swisstopo (light)") |>
  addProviderTiles("OpenStreetMap",     group = "OpenStreetMap") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addRasterImage(lf_dmr,  colors = pal_mr,  opacity = 0.8,
                 group = "Dist Main Road — Aargau [β = 0.244]") |>
  addRasterImage(lf_dsr,  colors = pal_sr,  opacity = 0.8,
                 group = "Dist Secondary Road — Aargau [β = 0.171]") |>
  addRasterImage(lf_dra,  colors = pal_rw,  opacity = 0.8,
                 group = "Dist Railway — Aargau [β = 0.505]") |>
  addRasterImage(lf_dse,  colors = pal_se,  opacity = 0.8,
                 group = "Dist Settlement — Aargau [β = 0.210]") |>
  addRasterImage(lf_dfr,  colors = pal_fr,  opacity = 0.8,
                 group = "Dist Fenced Road — Aargau [excluded]") |>
  addRasterImage(lf_dmr_zh, colors = pal_mr, opacity = 0.8,
                 group = "Dist Main Road — Zürich") |>
  addRasterImage(lf_dsr_zh, colors = pal_sr, opacity = 0.8,
                 group = "Dist Secondary Road — Zürich") |>
  addRasterImage(lf_dra_zh, colors = pal_rw, opacity = 0.8,
                 group = "Dist Railway — Zürich") |>
  addRasterImage(lf_dse_zh, colors = pal_se, opacity = 0.8,
                 group = "Dist Settlement — Zürich") |>
  addLayersControl(
    baseGroups    = base_groups,
    overlayGroups = c("Dist Main Road — Aargau [β = 0.244]",
                      "Dist Secondary Road — Aargau [β = 0.171]",
                      "Dist Railway — Aargau [β = 0.505]",
                      "Dist Settlement — Aargau [β = 0.210]",
                      "Dist Fenced Road — Aargau [excluded]",
                      "Dist Main Road — Zürich",
                      "Dist Secondary Road — Zürich",
                      "Dist Railway — Zürich",
                      "Dist Settlement — Zürich"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup(c("Dist Secondary Road — Aargau [β = 0.171]",
              "Dist Fenced Road — Aargau [excluded]",
              "Dist Main Road — Zürich",
              "Dist Secondary Road — Zürich",
              "Dist Railway — Zürich",
              "Dist Settlement — Zürich")) |>
  setView(lng = 8.15, lat = 47.38, zoom = 10) |>
  addLegend(pal = pal_mr, values = c(0, 7100),
            title = "Distance (m)", position = "bottomright")
```

### Barrier and Passage Layers

```{r}
#| label: map-barriers

pal_bin_red   <- colorFactor(c("#e74c3c"), domain = 1, na.color = "transparent")
pal_bin_blue  <- colorFactor(c("#2980b9"), domain = 1, na.color = "transparent")
pal_bin_orange<- colorFactor(c("#e67e22"), domain = 1, na.color = "transparent")
pal_bin_green <- colorFactor(c("#27ae60"), domain = 1, na.color = "transparent")

leaflet(options = leafletOptions(minZoom = 8)) |>
  addProviderTiles("CartoDB.Positron",  group = "Swisstopo (light)") |>
  addProviderTiles("OpenStreetMap",     group = "OpenStreetMap") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addRasterImage(lf_hw,   colors = pal_bin_red,    opacity = 0.9,
                 group = "Highway barriers — Aargau") |>
  addRasterImage(lf_ll,   colors = pal_bin_blue,   opacity = 0.9,
                 group = "Large lakes — Aargau") |>
  addRasterImage(lf_ls,   colors = pal_bin_orange, opacity = 0.7,
                 group = "Large settlements — Aargau") |>
  addRasterImage(lf_pa,   colors = pal_bin_green,  opacity = 0.9,
                 group = "Wildlife passages — Aargau") |>
  addRasterImage(lf_hw_zh,  colors = pal_bin_red,    opacity = 0.9,
                 group = "Highway barriers — Zürich") |>
  addRasterImage(lf_ll_zh,  colors = pal_bin_blue,   opacity = 0.9,
                 group = "Large lakes — Zürich") |>
  addRasterImage(lf_ls_zh,  colors = pal_bin_orange, opacity = 0.7,
                 group = "Large settlements — Zürich") |>
  addRasterImage(lf_pa_zh,  colors = pal_bin_green,  opacity = 0.9,
                 group = "Wildlife passages — Zürich") |>
  addLayersControl(
    baseGroups    = base_groups,
    overlayGroups = c("Highway barriers — Aargau",
                      "Large lakes — Aargau",
                      "Large settlements — Aargau",
                      "Wildlife passages — Aargau",
                      "Highway barriers — Zürich",
                      "Large lakes — Zürich",
                      "Large settlements — Zürich",
                      "Wildlife passages — Zürich"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup(c("Large lakes — Aargau",
              "Large settlements — Aargau",
              "Highway barriers — Zürich",
              "Large lakes — Zürich",
              "Large settlements — Zürich",
              "Wildlife passages — Zürich")) |>
  setView(lng = 8.3, lat = 47.40, zoom = 9) |>
  addLegend(colors = c("#e74c3c","#2980b9","#e67e22","#27ae60"),
            labels = c("Highway", "Large lake", "Large settlement", "Wildlife passage"),
            title = "Barrier / passage type", position = "bottomright",
            opacity = 0.85)
```

### Final Output Rasters

```{r}
#| label: map-outputs

leaflet(options = leafletOptions(minZoom = 8)) |>
  addProviderTiles("CartoDB.Positron",  group = "Swisstopo (light)") |>
  addProviderTiles("OpenStreetMap",     group = "OpenStreetMap") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addRasterImage(lf_suit_ag, colors = pal_suit, opacity = 0.85,
                 group = "Habitat Suitability — Aargau") |>
  addRasterImage(lf_res_ag,  colors = pal_res,  opacity = 0.85,
                 group = "Resistance — Aargau") |>
  addRasterImage(lf_suit_zh, colors = pal_suit, opacity = 0.85,
                 group = "Habitat Suitability — Zürich") |>
  addRasterImage(lf_res_zh,  colors = pal_res,  opacity = 0.85,
                 group = "Resistance — Zürich") |>
  addLayersControl(
    baseGroups    = base_groups,
    overlayGroups = c("Habitat Suitability — Aargau",
                      "Resistance — Aargau",
                      "Habitat Suitability — Zürich",
                      "Resistance — Zürich"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup(c("Habitat Suitability — Aargau",
              "Habitat Suitability — Zürich",
              "Resistance — Zürich")) |>
  setView(lng = 8.35, lat = 47.40, zoom = 9) |>
  addLegend(pal = pal_res, values = c(1, 55),
            title = "Resistance (R)", position = "bottomright") |>
  addLegend(pal = pal_suit, values = c(0, 1),
            title = "Suitability (S)", position = "bottomleft")
```
:::
::::